Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-03-03 jso.qmd
done with proofing

The second simulation examines various factors affecting the performance of JSO.
The JSO performance is evaluated under different hyper-parameters combinations
(number of jellyfish and the maximum number of tries). The effect of index properties (smoothness and squintability) along
with JSO hyper-parameters and data dimension variations is captured and studied using logistic regression.

@@

To examine the performance of JSO across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts.

In addition, to assess the performance of JSO across various PP problems, two different data shapes, pipe and a sine wave, are considered in various dimensional spaces using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. An example of holes and pipes data with the syntax to generate them are provided in the supplementary material. This results in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. This study does not consider dimensions lower than four, as exploring the search space in such cases is relatively straightforward. Similarly, extremely high-dimensional cases with complex data structures, where no method can reliably and consistenly uncover the underlying structure due to the curse of dimensionality, are also excluded, as they fall outside the scope of this study. A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability. Again, JSO is run 50 times for each scenario to calculate the success rate for each hyper-parameter combination.

Smoothness and squintability are computed following the procedures
outlined in @sec-smoothness and @sec-squintability and as illustrated in
@fig-smoothness and @fig-squintability. To compute smoothness, 500
random bases are simulated. Index values are calculated for each random
basis, followed by fitting a Gaussian process model to obtain the
smoothness measure for the index.

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases. A
binning procedure is applied to average the index values within each
0.005 projection distance bin. The four-parameter scaled logistic
function (@eq-parametric) is fitted to the index values against
projection distances, estimated by non-linear least squares to obtain
the squintability measure as Equation (@eq-squintability-parametric).

A generalised linear model is fitted using a binomial family and a logit link function to assess the factors affecting success rate.  Predictors are smoothness,
squintability, and JSO hyper-parameters. A success rate of 0 indicates the optimiser did not come sufficiently close to the optimal projection. <!-- The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine-wave shape.-->

@@

@@

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For problems with relatively lower dimensional search spaces (4 and 6 dimensions), small parameter values (20 jellyfish and a maximum of 50 tries) are sufficient for space exploration and thus can already achieve a high success rate. Higher parameter values (i.e. 100 jellyfish and a maximum of 100 tries) enhances exploration of the search space, which is particularly beneficial in higher-dimensional problems (8, 10, and 12 dimensions) where a more complex search is necessary. However, in lower-dimensional problems, the benefit may be less pronounced, as the space is already relatively easy to navigate. While increasing both parameters enhances the performance of JSO, it also extends the computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

Smoothness and squintability are calculated across a collection of
pipe-finding and sine-wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. @tbl-smoothness-squintability presents the parameters
estimated from the Gaussian process (outputscale $\eta$, lengthscale
$\ell$, smoothness $\nu$, and nugget $\sigma$) and from the scaled
logistic function ($\theta_1$ to $\theta_4$) for calculating smoothness
and squintability in each PP problem considered. The column $\nu$ is
used as the smoothness metric and the column $\varsigma$ is calculated
as equation (@eq-squintability-parametric) as the squintability metric.

@@

```{r}
#| label: fig-idx-proj-dist
#| fig.width: 8
#| fig.height: 4
#| fig.cap: "Five traces of index value against projection distance for each of the three indexes (TIC, spline, and skinny) in 6D sine data. The traces of the splines index are smooth, with a steep increaes before leveling off, suggesting the optimum can be detected from a far distance, hence a largest squintability value. In contrast, the skinny index (squintability of almost zeor) has a noisy trace that only stabilizes near the optimum, suggesting the optimizer needs get very close to detect it."
sq_basis_df_6d_TIC_spline_skinny |>
  ggplot(aes(x = dist, y = index, group = interaction(index_name, id), color = index_name)) +
  geom_line() + 
  scale_color_brewer(palette = "Dark2") +
  xlab("Projection distance") + ylab("Index value") +
  theme_bw() + 
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (outputscale $\eta$, lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and scaled logistic function ($\theta_1$ to $\theta_4$ and $\varsigma$) for the pipe-finding and sine-wave finding problems.  The columns $\nu$ and $\varsigma$ represent the smoothness and squintability measures respectively.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 17)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor", index),
         index = ifelse(index == "loess2d", "loess", index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  filter(shape == "sine", index %in% c("TIC", "splines", "skinny")) |>
  select(shape, index, d, smooth, squint) |>
  arrange(desc(squint)) |>
  knitr::kable(
    digits = 4, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      #"$\\eta$", "$\\ell$", 
      "$\\nu$", #"$\\sigma$",
      #"$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", 
      "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rr") #align = "|llr|rrrr|rrrrr") |> 
  #column_spec(1, border_left = TRUE) |> 
  #column_spec(c(4, 8, 13), border_right = TRUE) |> 
  #kable_styling(font_size = 10) 
```

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies. As expected, the JSO success rate is higher with higher squintability and/or more jellyfish, and lower when dimension is higher (higher-dimensional optimization is more difficult). Interestingly, the JSO is unaffected by the smoothness of the index function. This is consistent with the way random search algorithms jump from value to value without taking local regularity into account, as opposed to gradient-based optimizers for example. Allowing
for more tries also does not affect success rate significantly. A unit increase in squintability increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As dimension
increases by one the success rate almost halves. Increasing the number of
jellies by 10 increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.

@@

The simulation result suggests that squintability and smoothness are likely dependent. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness. While it may seem counterintuitive at first that smoothness is not a significant factor, this outcome is reasonable in this context, as the swarm-based method used in this study, JSO, is designed to perform well even in non-smooth search spaces by leveraging population-based exploration.

This study intends to serve as an illustration that the proposed measure effectively captures the key factors influencing the success rate of PP. The interactions between the two variables were not examined in this study due to the limited number of data points, nor was the linearity assumption of the relationship with success rate explicitly tested.

>>>>

@@

To examine the performance of JSO relative to hyper-parameter choices, the `holes` index is used to find the 2D pipe in data of dimensions 4-12. Additional variables are Gaussian noise. The levels of hyper-parameters are 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts. Fifty samples of each data set are generated, and the success rate is calculated in the same manner as described in @sec-app-1.

This set of simulations is expanded to assess the performance of JSO relative to the index properties, with a second data set type and additional PP indexes. The second data set has a sine wave in two dimensions and Gaussian noise in the remaining dimensions. Six additional PP indexes that should be able to extract the sine wave are used: `dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`, following @laa_using_2020. (The supplementary material includes definitions of these PP indexes, and descriptions of the pipe and sine wave data simulation.). The JSO hyper-parameters, number of jellies and maximum tries, are also included. This results in a total of `r nrow(sim_df)` scenarios, comprising of `r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and `r nrow(filter(sim_df, index != "holes"))` on the sine wave data. The choice of 4-12 dimensions covers the range of easy to relatively difficult for the optimisers to detect the structured projections. <!-- A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability.--> The same approach is used, that 50 samples are generated and the success rate is calculated at each level of the simulation design. A success rate of 0 indicates the optimiser did not come sufficiently close to the optimal projection.

@@

@fig-smoothness describes the procedure to compute smoothness. Index values are calculated for each of 500 simulated random bases. A Gaussian process model is fitted to the resulting data to obtain the smoothness measure for the index, as described in @sec-smoothness.

@@

@fig-squintability describes the procedure to compute squintability. Index values and
projection distances are computed for the projections along interpolated paths between 50 random bases and the optimal basis. An interpolation step size of 0.005 is used. The projection distances are binned to 0.005 intervals and index values are averaged in each bin. The four-parameter scaled logistic function (@eq-parametric) is fitted to the index values against projection distances, estimated by non-linear least squares to obtain the squintability measure as Equation (@eq-squintability-parametric).

A generalised linear model is fitted using a quasibinomial family and a logit link function to assess the factors affecting success rate.  Predictors are smoothness,
squintability, data dimension and JSO hyper-parameters. <!-- The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine wave shape.-->

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of tries) on the success rate is summarised in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For problems with relatively lower dimensional (4, 6) search spaces, small parameter values (20 jellyfish and a maximum of 50 tries) are sufficient to achieve a high success rate. Many more jellyfish (100 jellyfish and a maximum of 100 tries) are needed to obtain a mediocre success rate for dimensions 10 and 12. This suggests that even JSO will struggle with high dimensions, and require substantial more computational time to have an acceptable success rate, especially for difficult PP indexes based on scagnostics. <!-- dimensions) where a more complex search is necessary. However, in lower-dimensional problems, the benefit may be less pronounced, as the space is already relatively easy to navigate. While increasing both parameters enhances the performance of JSO, it also extends the computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.-->

@@

Smoothness and squintability are calculated across a collection of
pipe and sine wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. @tbl-smoothness-squintability presents values computed for a subset of PP indexes on the sine wave data for dimensions 4-8. (A full table of the parameters
estimated from the Gaussian process (outputscale $\eta$, lengthscale
$\ell$, smoothness $\nu$, and nugget $\sigma$) and from the scaled
logistic function ($\theta_1$ to $\theta_4$) for calculating smoothness
and squintability for all PP indexes and data combinations is included in the supplementary materials). The column $\nu$ is
used as the smoothness metric and the column $\varsigma$ is calculated
as equation (@eq-squintability-parametric) as the squintability metric.

@fig-idx-proj-dist shows a sample of traces from the indexes in @tbl-smoothness-squintability for the 6D sine wave data. The traces the PP index values computed for projections along an interpolated path between a noise projection and the optimal projection (sine wave). (Note index values were self-standardised to range between 0 and 1, for comparison.) The `splines` index is very smooth with a gradual increase in value between the noise projection and the optimal projection. The `TIC` index is slightly less smooth and slightly less gradual change in value. Least desirable is the `skinny` index which is noisier and the optimisation needs to be much closer to the optimal angle to have a higher value. This illustrates how the new measures for squintability and smoothness work to describe the PP indexes.

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: The columns $\nu$ and $\varsigma$ show the smoothness and squintability measures for three PP indexes for the sine wave data. The splines index has the largest squint angle and is the smoothest. The skinny index has low smoothness and smallest squint angle. As dimension increases the squint angle gets smaller but smoothness doesn't change much.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 17)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor", index),
         index = ifelse(index == "loess2d", "loess", index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  filter(shape == "sine", index %in% c("TIC", "splines", "skinny")) |>
  select(shape, index, d, smooth, squint) |>
  arrange(desc(squint)) |>
  knitr::kable(
    digits = 4, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      #"$\\eta$", "$\\ell$", 
      "$\\nu$", #"$\\sigma$",
      #"$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", 
      "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rr") #align = "|llr|rrrr|rrrrr") |> 
  #column_spec(1, border_left = TRUE) |> 
  #column_spec(c(4, 8, 13), border_right = TRUE) |> 
  #kable_styling(font_size = 10) 
```

@@

```{r}
#| label: fig-idx-proj-dist
#| fig.width: 8
#| fig.height: 4
#| fig.cap: "Five traces of index value against projection distance for each of the three indexes (TIC, spline, and skinny) on 6D sine wave data. The traces of the splines index are smooth, with a gradual change between noise and otimal projection. In contrast, the skinny index has a noisy trace that increases rapidly near the optimum. This illustrates the usefulness of the new measures for squintability and smoothness."
sq_basis_df_6d_TIC_spline_skinny |>
  ggplot(aes(x = dist, y = index, group = interaction(index_name, id), color = index_name)) +
  geom_line() + 
  scale_color_brewer(palette = "Dark2") +
  xlab("Projection distance") + ylab("Index value") +
  theme_bw() + 
  theme(legend.position = "bottom")
```

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies. As expected, the JSO success rate is higher with higher squintability and/or more jellyfish, and lower when dimension is higher (higher-dimensional optimization is more difficult). Interestingly, the JSO is unaffected by the smoothness of the index function. This is consistent with the way random search algorithms jump from value to value without taking local regularity into account, as opposed to gradient-based optimizers for example. Allowing
for more tries also does not affect success rate significantly. A unit increase in squintability increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As dimension
increases by one the success rate almost halves. Increasing the number of
jellies by 10 increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%. @tbl-joint-output shows the fit statistics supporting that the model is reasonably strong.

Note that the addition of interaction terms for squintability and data dimension, and smoothness and data dimension, improves the model a small amount, but does not change the interpretation above.

@@

<!--The simulation result suggests that squintability and smoothness are likely dependent. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness. While it may seem counterintuitive at first that smoothness is not a significant factor, this outcome is reasonable in this context, as the swarm-based method used in this study, JSO, is designed to perform well even in non-smooth search spaces by leveraging population-based exploration.-->

This study intends to serve as an illustration that the proposed measures effectively captures the key factors influencing the success rate of PP. <!--The interactions between the two variables were not examined in this study due to the limited number of data points, nor was the linearity assumption of the relationship with success rate explicitly tested.-->

### 2025-03-03 jso.tex
done with proofing

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, breakable, left=2mm, bottomtitle=1mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, rightrule=.15mm, toptitle=1mm, opacityback=0, bottomrule=.15mm, arc=.35mm, leftrule=.75mm]

@@

To examine the performance of JSO across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts.

In addition, to assess the performance of JSO across various PP
problems, two different data shapes, pipe and a sine wave, are
considered in various dimensional spaces using six different PP indexes:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, under varied JSO
hyper-parameters. An example of holes and pipes data with the syntax to
generate them are provided in the supplementary material. This results
in a total of 76 scenarios, comprising of 30 computed on the pipe data
and 46 on the sine-wave data. This study does not consider dimensions
lower than four, as exploring the search space in such cases is
relatively straightforward. Similarly, extremely high-dimensional cases
with complex data structures, where no method can reliably and
consistenly uncover the underlying structure due to the curse of
dimensionality, are also excluded, as they fall outside the scope of
this study. A list of combinations of the data shape, dimensions and PP
indexes that are considered in this study is provided in
Table~\ref{tbl-smoothness-squintability}. Again, JSO is run 50 times for
each scenario to calculate the success rate for each hyper-parameter
combination.

Smoothness and squintability are computed following the procedures
outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability} and as illustrated in
Figure~\ref{fig-smoothness} and Figure~\ref{fig-squintability}. To
compute smoothness, 500 random bases are simulated. Index values are
calculated for each random basis, followed by fitting a Gaussian process
model to obtain the smoothness measure for the index.

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases. A
binning procedure is applied to average the index values within each
0.005 projection distance bin. The four-parameter scaled logistic
function (\ref{eq-parametric}) is fitted to the index values against
projection distances, estimated by non-linear least squares to obtain
the squintability measure as Equation
(\ref{eq-squintability-parametric}).

A generalised linear model is fitted using a binomial family and a logit
link function to assess the factors affecting success rate. Predictors
are smoothness, squintability, and JSO hyper-parameters. A success rate
of 0 indicates the optimiser did not come sufficiently close to the
optimal projection.

@@

@@

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For problems
with relatively lower dimensional search spaces (4 and 6 dimensions),
small parameter values (20 jellyfish and a maximum of 50 tries) are
sufficient for space exploration and thus can already achieve a high
success rate. Higher parameter values (i.e.~100 jellyfish and a maximum
of 100 tries) enhances exploration of the search space, which is
particularly beneficial in higher-dimensional problems (8, 10, and 12
dimensions) where a more complex search is necessary. However, in
lower-dimensional problems, the benefit may be less pronounced, as the
space is already relatively easy to navigate. While increasing both
parameters enhances the performance of JSO, it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

Smoothness and squintability are calculated across a collection of
pipe-finding and sine-wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. Table~\ref{tbl-smoothness-squintability} presents the
parameters estimated from the Gaussian process (outputscale \(\eta\),
lengthscale \(\ell\), smoothness \(\nu\), and nugget \(\sigma\)) and
from the scaled logistic function (\(\theta_1\) to \(\theta_4\)) for
calculating smoothness and squintability in each PP problem considered.
The column \(\nu\) is used as the smoothness metric and the column
\(\varsigma\) is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

@@

\caption{\label{fig-idx-proj-dist}Five traces of index value against
projection distance for each of the three indexes (TIC, spline, and
skinny) in 6D sine data. The traces of the splines index are smooth,
with a steep increaes before leveling off, suggesting the optimum can be
detected from a far distance, hence a largest squintability value. In
contrast, the skinny index (squintability of almost zeor) has a noisy
trace that only stabilizes near the optimum, suggesting the optimizer
needs get very close to detect it.}

@@

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (outputscale \(\eta\), lengthscale \(\ell\),
smoothness \(\nu\), and nugget \(\sigma\)) and scaled logistic function
(\(\theta_1\) to \(\theta_4\) and \(\varsigma\)) for the pipe-finding
and sine-wave finding problems. The columns \(\nu\) and \(\varsigma\)
represent the smoothness and squintability measures respectively.}

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies.
As expected, the JSO success rate is higher with higher squintability
and/or more jellyfish, and lower when dimension is higher
(higher-dimensional optimization is more difficult). Interestingly, the
JSO is unaffected by the smoothness of the index function. This is
consistent with the way random search algorithms jump from value to
value without taking local regularity into account, as opposed to
gradient-based optimizers for example. Allowing for more tries also does
not affect success rate significantly. A unit increase in squintability
increases the success rate by 24\%. As dimension increases by one the
success rate almost halves. Increasing the number of jellies by 10
increases the success rate by 32\%.

\begingroup
\fontsize{12.0pt}{14.4pt}\selectfont
\setlength{\LTpost}{0mm}

@@

The simulation result suggests that squintability and smoothness are
likely dependent. Intuitively, a highly smooth search space exhibits
gradual changes, making it easier to perceive the overall structure from
a distance, thereby increasing the squint angle.
Table~\ref{tbl-joint-output} compare the model fits when squintability
and smoothness are included separately. Given that the null deviance
remains the same across models, the model with squintability yields a
smaller residual deviance and lower AIC and BIC values, indicating a
better fit. This suggests that squintability captures the relevant
variation more effectively than smoothness. While it may seem
counterintuitive at first that smoothness is not a significant factor,
this outcome is reasonable in this context, as the swarm-based method
used in this study, JSO, is designed to perform well even in non-smooth
search spaces by leveraging population-based exploration.

This study intends to serve as an illustration that the proposed measure
effectively captures the key factors influencing the success rate of PP.
The interactions between the two variables were not examined in this
study due to the limited number of data points, nor was the linearity
assumption of the relationship with success rate explicitly tested.

>>>>

\begin{tcolorbox}[enhanced jigsaw, colback=white, opacitybacktitle=0.6, breakable, bottomrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, opacityback=0, bottomtitle=1mm, coltitle=black, colframe=quarto-callout-note-color-frame, toptitle=1mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, left=2mm]

@@

To examine the performance of JSO relative to hyper-parameter choices,
the \texttt{holes} index is used to find the 2D pipe in data of
dimensions 4-12. Additional variables are Gaussian noise. The levels of
hyper-parameters are 20, 50, and 100 jellyfish and a maximum of 50 and
100 attempts. Fifty samples of each data set are generated, and the
success rate is calculated in the same manner as described in
Section~\ref{sec-app-1}.

This set of simulations is expanded to assess the performance of JSO
relative to the index properties, with a second data set type and
additional PP indexes. The second data set has a sine wave in two
dimensions and Gaussian noise in the remaining dimensions. Six
additional PP indexes that should be able to extract the sine wave are
used: \texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, following Laa and Cook (2020).
(The supplementary material includes definitions of these PP indexes,
and descriptions of the pipe and sine wave data simulation.). The JSO
hyper-parameters, number of jellies and maximum tries, are also
included. This results in a total of 76 scenarios, comprising of 30
computed on the pipe data and 46 on the sine wave data. The choice of
4-12 dimensions covers the range of easy to relatively difficult for the
optimisers to detect the structured projections. The same approach is
used, that 50 samples are generated and the success rate is calculated
at each level of the simulation design. A success rate of 0 indicates
the optimiser did not come sufficiently close to the optimal projection.

@@

Figure~\ref{fig-smoothness} describes the procedure to compute
smoothness. Index values are calculated for each of 500 simulated random
bases. A Gaussian process model is fitted to the resulting data to
obtain the smoothness measure for the index, as described in
Section~\ref{sec-smoothness}.

@@

Figure~\ref{fig-squintability} describes the procedure to compute
squintability. Index values and projection distances are computed for
the projections along interpolated paths between 50 random bases and the
optimal basis. An interpolation step size of 0.005 is used. The
projection distances are binned to 0.005 intervals and index values are
averaged in each bin. The four-parameter scaled logistic function
(\ref{eq-parametric}) is fitted to the index values against projection
distances, estimated by non-linear least squares to obtain the
squintability measure as Equation (\ref{eq-squintability-parametric}).

A generalised linear model is fitted using a quasibinomial family and a
logit link function to assess the factors affecting success rate.
Predictors are smoothness, squintability, data dimension and JSO
hyper-parameters.

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of tries) on the success rate is summarised in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For problems
with relatively lower dimensional (4, 6) search spaces, small parameter
values (20 jellyfish and a maximum of 50 tries) are sufficient to
achieve a high success rate. Many more jellyfish (100 jellyfish and a
maximum of 100 tries) are needed to obtain a mediocre success rate for
dimensions 10 and 12. This suggests that even JSO will struggle with
high dimensions, and require substantial more computational time to have
an acceptable success rate, especially for difficult PP indexes based on
scagnostics.

@@

Smoothness and squintability are calculated across a collection of pipe
and sine wave finding problems to construct the relationship between
success rate, JSO hyper-parameters, and index properties.
Table~\ref{tbl-smoothness-squintability} presents values computed for a
subset of PP indexes on the sine wave data for dimensions 4-8. (A full
table of the parameters estimated from the Gaussian process (outputscale
\(\eta\), lengthscale \(\ell\), smoothness \(\nu\), and nugget
\(\sigma\)) and from the scaled logistic function (\(\theta_1\) to
\(\theta_4\)) for calculating smoothness and squintability for all PP
indexes and data combinations is included in the supplementary
materials). The column \(\nu\) is used as the smoothness metric and the
column \(\varsigma\) is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

Figure~\ref{fig-idx-proj-dist} shows a sample of traces from the indexes
in Table~\ref{tbl-smoothness-squintability} for the 6D sine wave data.
The traces the PP index values computed for projections along an
interpolated path between a noise projection and the optimal projection
(sine wave). (Note index values were self-standardised to range between
0 and 1, for comparison.) The \texttt{splines} index is very smooth with
a gradual increase in value between the noise projection and the optimal
projection. The \texttt{TIC} index is slightly less smooth and slightly
less gradual change in value. Least desirable is the \texttt{skinny}
index which is noisier and the optimisation needs to be much closer to
the optimal angle to have a higher value. This illustrates how the new
measures for squintability and smoothness work to describe the PP
indexes.

\caption{\label{tbl-smoothness-squintability}The columns \(\nu\) and
\(\varsigma\) show the smoothness and squintability measures for three
PP indexes for the sine wave data. The splines index has the largest
squint angle and is the smoothest. The skinny index has low smoothness
and smallest squint angle. As dimension increases the squint angle gets
smaller but smoothness doesn't change much.}

@@

\caption{\label{fig-idx-proj-dist}Five traces of index value against
projection distance for each of the three indexes (TIC, spline, and
skinny) on 6D sine wave data. The traces of the splines index are
smooth, with a gradual change between noise and otimal projection. In
contrast, the skinny index has a noisy trace that increases rapidly near
the optimum. This illustrates the usefulness of the new measures for
squintability and smoothness.}

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies.
As expected, the JSO success rate is higher with higher squintability
and/or more jellyfish, and lower when dimension is higher
(higher-dimensional optimization is more difficult). Interestingly, the
JSO is unaffected by the smoothness of the index function. This is
consistent with the way random search algorithms jump from value to
value without taking local regularity into account, as opposed to
gradient-based optimizers for example. Allowing for more tries also does
not affect success rate significantly. A unit increase in squintability
increases the success rate by 24\%. As dimension increases by one the
success rate almost halves. Increasing the number of jellies by 10
increases the success rate by 32\%. Table~\ref{tbl-joint-output} shows
the fit statistics supporting that the model is reasonably strong.

Note that the addition of interaction terms for squintability and data
dimension, and smoothness and data dimension, improves the model a small
amount, but does not change the interpretation above.

\setlength{\LTpost}{0mm}

@@

This study intends to serve as an illustration that the proposed
measures effectively captures the key factors influencing the success
rate of PP.
